\myTitle

### 2024-10-02 tex/body-text-of-paper.tex
Clean up discussion

In the following, we present a new reference baseline that includes 1,636 Chinook salmon individuals
from 17 collections, genotyped at 204 microhaplotypes \citep{baetscher2018microhaplotypes} from throughout the genome.
This baseline is
highly effective for GSI within California and the markers within it provide ample power
for PBT.
We describe how we identified and compiled the markers, including several new markers
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that includes multiallelic gene regions with high variability for
relationship inference and gene regions identified in whole-genome sequence data
for increased power for population identification. We show how this set of markers
provides sufficient power for highly accurate identification of parent and offspring
pairs and full siblings in California Chinook salmon, and also allows near-perfect
identification of individuals to population or genetic group of origin. This set of
genetic markers lays the foundation for a comprehensive genetic monitoring and evaluation effort
that facilitates multiple types of inference and is flexible and extensible.

When, and if, sequencing costs drop sufficiently, it may be possible to include
all different populations of a single species within a single standardized reference baseline
that performs equally well at broad and regional scales. A statistical approach for population
assignment from low coverage whole genome sequences appeared recently
\citep{desaixINPRESSpopulation}, and its authors noted that such a data type could
be ideal for producing reference baselines simultaneously applicable to broad scale
coverage and high resolution within sub-regions.   For the present, however,
baselines tailored to specific regions are essential for regional management questions.

\comment{Maybe discuss: A large fraction of the Feather River
Hatchery Spring-run (FRHS) collection into a cluster (colored orange in Fig.~\ref{fig:struct}).
A large fraction of the samples from the fall-run hatchery
program at the Feather River (FRHF) are also members of that same orange cluster, while others
belong to a cluster composed of other fall-run Chinook salmon. These STRUCTURE results at $K=7$,
validate the finding that all populations outside of the }

Although previous work using genome-scale data has shown that many of
these populations or stocks can be resolved \cite{thompson2020complex,meek2020identifying},
such data are not practical for most large-scale applications of GSI and PBT, particularly because
of the much higher costs, longer turnaround time and challenges of producing highly
replicable datasets with reduced representation and low-coverage whole genome sequencing methods \cite{ali2016rad}.

Accurate identification of the distinct ecotypes of Chinook salmon in the
California Central Valley is of utmost importance for designing and implementing
effective conservation and management actions. This has been challenging,
given the recent common ancestors of these ecotypes and ongoing migration between subbasins.
We describe the first replicable, reproducible panel of genetic markers that identifies the late-fall-run salmon ecotype,
which occurs only in the CCV and shares a genomic background with the more
common CCV fall-run salmon ecotype. Moreover, we show how the FRH spring-run lineage is
easily identifiable through a combination of traditional GSI and the characterization
of functional genetic markers in the RoSA. Finally, although previous work has described
GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from
each other and their fall-run counterparts with moderately high accuracy, we demonstrate
near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are
apparently misidentified (Fig. 3) are likely to be primarily migrants and not true misidentifications.
For example, the three fish that were field-characterized as spring-run from Butte Creek, but are
genetically identified as winter-run salmon clearly carry the winter-run genomic background and
are strays from the winter-run stock, as these 
could not realistically be misidentified on the basis of the genetic data.

This marker set and baseline reference dataset also provides
unprecedented power for identifying fish from not only the other ESUs in California,
but even the individual populations within those ESUs (Fig. 3).

Previous genetic marker sets for California Chinook salmon had sufficient power for inferring
parent-offspring relationships, but only when both parents were sampled and genotyped.
This marker set considerably increases the capacity for relationship inference in California
salmon, by providing sufficient power for parent-offspring relationship inference, when only
one parent is sampled, as well as identifying pairs of full siblings when no parents are sampled.
Groups of full siblings larger than two are identified even more accurately than expressed in Fig.~\ref{fig:fprs}.
Although some additional errors might be
expected when related but non-target kin pairs are sampled,  the false positive rates
for these non-targe kin pairs are sufficiently small
for almost all realistic scenarios.
Accurate relationship
inference in even the most genetically depauperate Chinook salmon population in California
and rangewide \citep{seeb2007development,clemento2014evaluation} is therefore possible with this marker set.
We note that, as in previous work, it is nearly impossible to identify HS pairs accurately
with data from this, or any standard marker dataset. Accurate identification of HS pairs
typically relies on thousands of SNP markers \citep{hillary2018genetic}.

Results from the model-based clustering analysis with the program STRUCTURE, revealed patterns of population relationships that are coincident with previous work (Clemento et al. 2014; Kinziger et al. 2013), emphasizing the distinction between Chinook salmon
populations in Coastal California and the Central Valley. These analyses also uncover some additional patterns, including the clear
presence of mixed ancestry in the Coastal California population in the Russian River, which is consistent with its location, at the southernmost edge of the Coastal Chinook salmon between the mouths of the Sacramento River and the other Coastal Chinook salmon basins to the north.

>>>>

In the following, we present a new reference baseline that includes 1,636 Chinook salmon individuals
from 17 collections, genotyped at 204 microhaplotypes \citep{baetscher2018microhaplotypes} from throughout the genome.
This baseline is
highly effective for GSI within California and the markers within it provide ample power
for PBT, thus enabling a highly effective and efficient integrated GSI/PBT monitoring and evaluation program 
\citep{garza2007large,beacham2021parentage}.
We describe how we identified and compiled the markers, including several new markers
that are particularly effective at distinguishing between closely related groups of fish,
such as late-fall and fall-run Chinook salmon in the CCV.  We also provide a comprehensive analysis of this
set of markers for inference of parents and siblings within different groups of populations.

@@

We describe a novel panel of microhaplotype genetic markers for Chinook salmon
in California that provides sufficient power for highly accurate identification 
of parent and offspring pairs and full siblings, and also allows near-perfect
identification of individuals to population or genetic group of origin in California. 
This set of genetic markers includes multiallelic gene regions with high variability for
relationship inference, gene regions identified in whole-genome sequence data
for increased power for identification of specific populations and incorporates markers that have
long been in use for both GSI and PBT in California (Clemento et al. 2014), by converting them
into so-called 'SNPlicons'. This set of markers lays the foundation for a comprehensive 
genetic monitoring and evaluation effort that facilitates multiple types of inference 
and is flexible and extensible.

This marker set and baseline reference dataset provide
unprecedented power for identifying fish from all of the Chinook salmon ESUs in California,
as well as individual populations within those ESUs (Fig.~\ref{fig:gsi}).
Accurate identification of the distinct ecotypes of Chinook salmon in the
California Central Valley is of utmost importance for designing and implementing
effective conservation and management actions. The CCV is one of the largest river
basins on the west coast of North America and has the highest diversity of recognized Chinook salmon
ecotypes in the species range, with four named ecotypes, two of which are protected
under the US Endangered Species Act. This basin drains the Sierra Nevada and southern Cascade
mountain ranges and is the source of water for several major urban areas as well as the
largest and most valuable agriculture area in North America. As such, monitoring and
evaluation of individual ecotypes and populations is of critical importance. This has been challenging,
given the recent common ancestors of these ecotypes \citep{clemento2014evaluation}, and
ongoing migration between subbasins where different ecotypes predominate.

We describe the first set of genetic markers that produces easily replicable data and
that identifies all of the ecotypes in the CCV, including the late-fall-run Chinook salmon ecotype.
The late-fall-run occurs only in the CCV and shares a genomic background with the more
common CCV fall-run salmon ecotype, so has been refractory to previous GSI efforts with other marker types \citep{seeb2007development,clemento2014evaluation,thompson2024genomics}. Moreover, we show how the FRH spring-run lineage is
easily identifiable through a combination of traditional GSI and the characterization
of functional genetic markers in the RoSA. Finally, although previous work has described
GSI capabilities that distinguish the natural-spawning CCV spring-run lineages from
each other and their fall-run counterparts with moderately high accuracy, we demonstrate
near complete accuracy in distinguishing these 'stocks'. Moreover, the few fish that are
apparently misidentified (Fig. 3) are likely to be primarily migrants and not true misidentifications.
For example, the three fish that were field-characterized as spring-run from Butte Creek, but are
genetically identified as winter-run salmon clearly carry the winter-run genomic background and
are strays from the winter-run stock, as these 
could not realistically be misidentified on the basis of the genetic data.

As sequencing costs drop, whole genome sequence (WGS) data may become the preferred
data type for salmon genetics. If so, it may become possible to include
all >200 genetically distinct Chinook salmon populations within a single standardized reference baseline
constructed with WGS data
that performs equally well at broad and regional scales \citep{desaixINPRESSpopulation}.
Such an approach would be highly flexible and extensible, as it would allow for the use
just about any marker type, including reduced representation DNA sequencing (e.g. RADseq,
\citealt{meek2020identifying,thompson2024genomics}),
as regional monitoring datasets would not need to be replicable.  For the present, however,
baselines tailored to specific regions are essential for regional management questions and
targeted sequencing approaches have proven to be the most practical for large-scale
applications of either GSI, PBT or an integrated monitoring program \citep{beacham2021parentage}.

%\comment{Maybe discuss: A large fraction of the Feather River
%Hatchery Spring-run (FRHS) collection into a cluster (colored orange in Fig.~\ref{fig:struct}).
%A large fraction of the samples from the fall-run hatchery
%program at the Feather River (FRHF) are also members of that same orange cluster, while others
%belong to a cluster composed of other fall-run Chinook salmon. These STRUCTURE results at $K=7$,
%validate the finding that all populations outside of the }

%Although previous work using genome-scale data has shown that many of
%these populations or stocks can be resolved \cite{thompson2020complex,meek2020identifying},
%such data are not practical for most large-scale applications of GSI and PBT, particularly because
%of the much higher costs, longer turnaround time and challenges of producing highly
%replicable datasets with reduced representation and low-coverage whole genome sequencing methods \cite{ali2016rad}.

Previous genetic marker sets for California Chinook salmon had sufficient power for inferring
parent-offspring relationships, but only when both parents were sampled and genotyped.
This marker set considerably increases the capacity for relationship inference in California
salmon, by providing sufficient power for parent-offspring relationship inference, when only
one parent is sampled, as well as identifying pairs of full siblings when no parents are sampled.
Groups of full siblings larger than two are identified even more accurately than expressed in Fig.~\ref{fig:fprs}.
Although some additional errors might be
expected when related but non-target kin pairs are sampled,  the false positive rates
for these non-targe kin pairs are sufficiently small
for almost all realistic scenarios.
Accurate relationship
inference in even the most genetically depauperate Chinook salmon population in California
and rangewide \citep{seeb2007development,clemento2014evaluation} is therefore possible with this marker set.
We note that, as in previous work, it is nearly impossible to identify HS pairs accurately
with data from this, or any standard marker dataset. Accurate identification of HS pairs
typically relies on many hundreds of microhaplotype markers \citep{baetscher2018microhaplotypes} or thousands of SNP markers \citep{hillary2018genetic}.

Results from the model-based clustering analysis with the program STRUCTURE, revealed patterns of population relationships that are
coincident with previous work (Clemento et al. 2014; Kinziger et al. 2013), emphasizing the distinction between Chinook salmon
populations in Coastal California and the Central Valley. These analyses also uncover some additional patterns, including the clear
presence of mixed ancestry in the Coastal California population in the Russian River, which is consistent with its location, at the 
southernmost edge of the Coastal Chinook salmon distribution and proximate to the mouth of the Sacramento River (the Golden Gate). In the
Central Valley, the clustering results emphasize the genetic distinction of the SWR population, which is likely at least partially due to 
the extreme bottleneck that it passed through 15-20 generations prior. The inclusion of the LFAR markers also resulted in the clear 
distinction of the CHLF group, which is coincident with the long-known phenotypic distinction of this population, but represents a novel 
genetic result. Moreover, at K = 7, the fall-run reporting group breaks into two clusters, which are distributed across almost all of the 
spring and fall-run populations in the Central Valley, albeit not equivalently, emphasizing both current and historical  migration and 
gene flow between them.
